OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2017-06-02 main.tex
Wrote the introduction.

\abstract{
  \textbf{Motivation:} The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented advance in our understanding of genome
structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but none of them addresses the common issue of
abnormal karyotypes. Many experiments are performed in cancer cell lines,
which typically harbor large-scale copy number variations that create
visible defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps have not been investigated so far.\\
	\textbf{Results:}
We observed that current normalization methods perform badly in the
presence of large-scale copy number variations, obscuring biological
viarations and enhancing batch effects. To address this issue, we
developed an alternative appoach designed to take into account chromosomal
abnormalities. The method, called OneD, increases reproducibility among
replicates of Hi-C samples with abnormal karyotype, thus significantly
outperforming previous methods.  In the absence of chromosomal
aberrantions, OneD fared equally well as state-of-the-art methods, making
it a safe choice for Hi-C normalization. OneD is fast and scales well in
terms of computing resources for resolutions up to 1 kbp.\\	
	\textbf{Availability:} OneD is implemented as an R package available at \href{http://www.github.com/qenvio/dryhic}{http://www.github.com/qenvio/dryhic}.\\	
	\textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
	\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}	online.}
\maketitle

@@

One of the crown achievements of modern biology was to realize that
genomes have an underlying three-dimensional structure contributing to
their activity (CITE). In mammals, this organization plays a key role in
guiding enhancer-promoter contacts (CITE), in V(D)J recombination (CITE)
and in X chromosome inactivation (CITE). One of the most significant
breakthroughs towards this insight was the development of the high
throughput chromosomal conformation capture (Hi-C) technology, assaying
chromosomal contacts at a genome-wide scale
\citep{lieberman2009comprehensive}. Exploring the spatial organization of
chromatin inside the cell is now common and Hi-C is part of the standard
molecular biology toolbox (CITE).

Contrary to the precursor technologies 3C, 4C and 5C \citep{de2012decade},
Hi-C interrogates all possible pairwise interactions between restriction
fragments. However, this does not guarantee the absence of bias. For
instance, local genome features such as the G+C content, the availability
of restriction enzyme sites and the mappability of the sequencing reads
have been shown to modify the results \citep{yaffe2011probabilistic}. In
addition, the general experimental biases such as batch effects and
protocol variation also apply. It is thus important to normalize
Hi-C data in order to remove biases and artifacts, so that they are not
confused with biological signal.

Several methods have been proposed to remove biases in Hi-C experiments
\citep{schmitt2016genome}. The first strategy is to model biases
explicitly from a defined set of local genomic features, such as the G+C
content and the availability of restriction sites. This approach is used
in the method of \cite{yaffe2011probabilistic} and in Hicnorm by
\cite{hu2012hicnorm}. The second strategy is to implicitly correct for
unknown biases by enforcing some regularity on the data. This approach is
used in the ICE method of \cite{imakaev2012iterative}, where the total
amount of contacts is imposed to be equal for every bin. ICE is currently
the most popular method, which is in part due to the speed of the
available implementation.

Both strategies are suitable for cells with a normal karyotype, but they
were not designed for cell types with karyotypic aberrations.

Although there have been some proposals that consider aneuploidy
\citep{wu2016computational}, to our knowledge, there is a lack of
methodology dealing with aberrant karyotypes (common in cancer cell lines)
that present abnormalities more complex than entire chromosome
duplications.

Here we propose \textit{oned}, an alternative designed to model total
number of contacts per genomic loci, thus taking into account possible
chromosomal abnormalities. It explicitly models the contribution of known
biases via a generalized additive model (GAM). The resulting data present
better overall reproducibility between replicas and across different
protocols, specially, but not restricted to, in the presence of aberrant
karyotypes. Besides, the implementation is fast and scales up to 1 Kbp
resolution with reasonable computing resources.

>>>>

\abstract{
  \textbf{Motivation:} The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented advance in our understanding of genome
structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but none of them addresses the common issue of
abnormal karyotypes. Many experiments are performed in cancer cell lines,
which typically harbor large-scale copy number variations that create
visible defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.\\
	\textbf{Results:}
We observed that current normalization methods perform badly in the
presence of large-scale copy number variations, obscuring biological
viarations and enhancing batch effects. To address this issue, we
developed an alternative appoach designed to take into account chromosomal
abnormalities. The method, called OneD, increases reproducibility among
replicates of Hi-C samples with abnormal karyotype, thus significantly
outperforming previous methods.  In the absence of chromosomal
aberrantions, OneD fared equally well as state-of-the-art methods, making
it a safe choice for Hi-C normalization. OneD is fast and scales well in
terms of computing resources for resolutions up to 1 kbp.\\	
	\textbf{Availability:} OneD is implemented as an R package available at \href{http://www.github.com/qenvio/dryhic}{http://www.github.com/qenvio/dryhic}.\\	
	\textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
	\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}	online.}
\maketitle

@@

One of the crown achievements of modern biology was to realize that
genomes have an underlying three-dimensional structure contributing to
their activity \citep{rowley2016three}. In mammals, this organization
plays a key role in guiding enhancer-promoter contacts (CITE), in V(D)J
recombination (CITE) and in X chromosome inactivation (CITE). One of the
most significant breakthroughs towards this insight was the development of
the high throughput chromosomal conformation capture technology (Hi-C),
assaying chromosomal contacts at a genome-wide scale
\citep{lieberman2009comprehensive}. Nowadays, exploring the spatial
organization of chromatin has become a priority in many fields and
Hi-C has become part of the standard molecular biology toolbox (CITE).

Contrary to the precursor technologies 3C, 4C and 5C \citep{de2012decade},
Hi-C interrogates all possible pairwise interactions between restriction
fragments. However, this does not guarantee that the method has no bias.
On the contrary, local genome features such as the G+C content, the
availability of restriction enzyme sites and the mappability of the
sequencing reads have been shown to impact the results
\citep{yaffe2011probabilistic}. In addition, the general experimental
biases such as batch effects and protocol variation also apply. It is thus
important to normalize Hi-C data in order to remove biases and artifacts,
so that they are not confused with biological signal.

Several methods have been proposed to remove biases in Hi-C experiments
\citep{schmitt2016genome}. The first strategy is to model biases
explicitly from a defined set of local genomic features, such as the G+C
content and the availability of restriction sites. This approach is used
in the method of \cite{yaffe2011probabilistic} and in Hicnorm by
\cite{hu2012hicnorm}. The second strategy is to implicitly correct
unknown biases by enforcing some regularity on the data. This approach is
used in the ICE method of \cite{imakaev2012iterative}, where the total
amount of contacts of every bin is imposed to be the same. ICE is
currently the most popular method, due in part to its speed.

Neither of these strategies were designed with regard for cell types with
karyotypic aberrations, most common in cancer cells in culture.  Hi-C is
very sensitive to aneuploidy, copy number variations and translocations.
Actually, these aberrations have so much influence on the outcome that the
artifacts can be used to re-assemble the target genome
\citep{korbel2013genome}.  So far the only attempt to address the issue
was the correction method caICB by \cite{wu2016computational}. However,
caICB applies a chromosome-wide correction, effectively excluding the
numerous cases of partial aneuploidy and regional copy number variations.

Here we propose \textit{OneD}, a method to correct local chromosomal
abnormalities. OneD explicitly models the contribution of known biases via
a generalized additive model. The normalized data is more reproducible
between replicates and across different protocols. Importantly, OneD is
also applicable when cells have a normal karyotype, where it performs as
well as the best normalization methods. Finally, the implementation is
faster than ICE and scales up to 1 kbp resolution with reasonable
computing resources.
